---
title: "Plots & Figures"
author:
  - Bénédicte Colnet [Inria, Paris-Saclay]
date: "June 2022"
output:
  html_document:
    code_folding: "hide"
    number_sections: no
    toc: yes
    toc_depth: 2
  pdf_document:
    toc: yes
abstract: | 
  This script reproduces all the plots from the paper entitled "Reweighting the RCT for generalization: finite sample analysis and variable selection". For simulatiosn, data are generated with separate scripts and outputs are stored in the folder entitled "results". This notebook only reads these results except for small additional illustrative Figures or simulations.
---
 
```{r setup, include=T}
knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

# libraries
library(MASS) # simulations
library(ggplot2) # plots
library(tidyr) # pivot longer or wider
library(dplyr) # case_when and others
library(ggridges)

source("./estimators.R")
source("./generateDGPs.R")

options(dplyr.summarise.inform = FALSE)

TARGET.ATE =  0.3*10+0.7*5
```


# Data generative process visualization


```{r}
N.for.plot = 15000

p.Y0 <- rnorm(n = N.for.plot, mean = 0, sd = 1)
p.Y1.X1 <- rnorm(n = N.for.plot, mean = 10, sd = 1)
p.Y1.X0 <- rnorm(n = N.for.plot, mean = 5, sd = 1)

DGP.for.plot <- data.frame("Y" = c(p.Y0, p.Y1.X1, p.Y1.X0),
                           "DGP" = c(rep("any X, Y(0)", N.for.plot), 
                                     rep("X = 1, Y(1)", N.for.plot), 
                                     rep("X = 0, Y(1)", N.for.plot)))
```


```{r}
ggplot(DGP.for.plot, aes(x = Y, y = DGP, group = DGP, fill = DGP)) + 
  geom_density_ridges(alpha = 0.6) +
  theme_classic() + 
  scale_fill_manual(values=c("#A2A399", "#149AD8", "#E69F00"), name = "") +
  xlab("Y") +
  ylab("Density") +
  theme(legend.position="none") +
  #theme(legend.position = c(0.82, 0.274)) +
  #theme( legend.text = element_text(size = 8, face = "bold")) +
  annotate(geom="text", x=15, y=1.3, label='atop(bold("Y(0)"))', size = 5, parse = TRUE) +
  annotate(geom="text", x=15, y=2.4, label='atop(bold("Y(1) | X = 0"))', size = 5, color = "darkblue", parse = TRUE) +
  annotate(geom="text", x=15, y=3.4, label='atop(bold("Y(1) | X = 1"))', size = 5, color = "darkorange", parse = TRUE) +
  xlim(-2, 17) +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) 

ggsave("./fig/toy_example_response_level.pdf", width = 4, height = 5)
```

```{r}
simulated.RCT <- toy.example(n = 150, m = 100, noisier.var.X1 = T, factor.var = 2)
```

```{r}
# focus on trial
simulated.RCT <- simulated.RCT[simulated.RCT$S == 1,]

# overall effect
estimated.ATE <- difference.in.means(simulated.RCT)
estimated.ATE.CI <- t.test(simulated.RCT[simulated.RCT$A == 1, "Y"], simulated.RCT[simulated.RCT$A == 0, "Y"])$conf.int

# X = 1
estimated.ATE.X.1 <- difference.in.means(simulated.RCT[simulated.RCT$X == 1,])
estimated.ATE.X.1.CI <- t.test(simulated.RCT[simulated.RCT$A == 1 & simulated.RCT$X == 1, "Y"], simulated.RCT[simulated.RCT$A == 0 & simulated.RCT$X == 1, "Y"])$conf.int

# X = 0
estimated.ATE.X.0 <- difference.in.means(simulated.RCT[simulated.RCT$X == 0,])
estimated.ATE.X.0.CI <- t.test(simulated.RCT[simulated.RCT$A == 1 & simulated.RCT$X == 0, "Y"], simulated.RCT[simulated.RCT$A == 0 & simulated.RCT$X == 0, "Y"])$conf.int


ate.stratified <- data.frame("ATE" = c(estimated.ATE, estimated.ATE.X.1, estimated.ATE.X.0),
                             "conf.inf" = c(estimated.ATE.CI[[1]], estimated.ATE.X.1.CI[[1]], estimated.ATE.X.0.CI[[1]]),
                             "conf.sup" = c(estimated.ATE.CI[[2]], estimated.ATE.X.1.CI[[2]], estimated.ATE.X.0.CI[[2]]),
                             "Strata" = c("All individuals", "X=1", "X=0"))
```

```{r}
ggplot(ate.stratified, aes(x = Strata, y = ATE, color = Strata)) +
  geom_point(size = 3.5) +
  geom_errorbar(aes(ymin=conf.inf, ymax=conf.sup), width=.2,
                 position=position_dodge(0.05), size = 1.5) +
  theme_minimal() + 
  #coord_flip() +
  xlab("") +
  ylab("Estimated effect") +
  theme(axis.text.x = element_text(size= 14, face="bold", color = "black"),
        axis.text.y = element_text(size= 11, face="bold", color = "black"),
        legend.position = "none") + 
  scale_color_manual(values=c("#222027",  "#56B4E9", "#E69F00")) 
ggsave("./fig/toy-example-stratification.png", width = 4, height = 3)
```


# Simple simulation to illustrate it works and compare $\pi$ and $\hat \pi$


```{r}
toy.example.pi.hat.or.not <- data.frame("estimate" = c(),
                                        "estimator" = c(),
                                        "estimating.pi" = c())


for (i in 1:1000){
  
  # generate data
  simulation <- toy.example(n = 150, m = 1000, output.oracles = F, symetric.po = F, noisier.var.X1 = F, factor.var = 0)
  
  ipsw.pi.oracle <- ipsw.univariate.and.categorical.X(dataframe = simulation, estimand = "ATE", oracle.e = T, oracle.pt = F, oracle.pr = F)
  
  ipsw.pi.hat <- ipsw.univariate.and.categorical.X(dataframe = simulation, estimand = "ATE", oracle.e = F, oracle.pt = F, oracle.pr = F)
  
  ht <- difference.in.means(dataframe = simulation, estimand = "ATE", oracle.e.too = T)
  
  dm  <- difference.in.means(dataframe = simulation, estimand = "ATE", oracle.e.too = F)
  
  
  new.row <- data.frame("estimate" = c(ipsw.pi.oracle, ipsw.pi.hat, ht, dm),
                        "estimator" = c("Re-weighted RCT \n (Equation 2) \n ", "Re-weighted RCT \n (Def. 9) \n " , "Horvitz-Thomson \n (Equation 1) \n ", "Difference-in-means"),
                        "estimating.pi" = c("Oracle pi","Estimated pi","Oracle pi","Estimated pi"))
  
  toy.example.pi.hat.or.not <- rbind(toy.example.pi.hat.or.not, new.row)
}
```

```{r}
toy.example.pi.hat.or.not$estimator <- factor(toy.example.pi.hat.or.not$estimator , levels = c("Horvitz-Thomson \n (Equation 1) \n ", "Difference-in-means", "Re-weighted RCT \n (Equation 2) \n ", "Re-weighted RCT \n (Def. 9) \n "))
```


```{r}
ggplot(toy.example.pi.hat.or.not[toy.example.pi.hat.or.not$estimator != "Difference-in-means", ], aes(x = estimator, y = estimate, fill = estimator)) +
  geom_boxplot(alpha = 0.8) +
  #geom_hline(aes(yintercept = 0.3*10+0.7*5, linetype = "ATE in Target population"),  size = 1, linetype = "dashed", color = "#EA471F",  show.legend =TRUE) + 
 # geom_hline(aes(yintercept = 0.75*10+0.25*5, linetype = "ATE in Trial population"), size = 1, linetype="dashed", color = "#475E96",  show.legend =TRUE)  +
  theme_minimal() +
  scale_fill_manual(values=c("#999999","#E69F00", "#EC420B")) +
  geom_hline(aes(yintercept = TARGET.ATE, color = "Population ATE"),  
               size = 0.9, linetype="dashed") +
  geom_hline(aes(yintercept = 0.75*10+0.25*5, color = "Trial ATE"), 
               size = 0.9, linetype="dashed") +
    xlab("") +
    ylab("Estimated ATE")  +
    theme(legend.title = element_blank(), legend.text = element_text(size=13, face = "bold"), legend.position = "right") +  
    theme(axis.text = element_text(angle = 0, size=13, face = "bold"))
ggsave("./fig/toy_example_simplest_expe.pdf", width = 9, height = 5)
```

# Effect of $n$ and $m$

```{r}
toy.example.n <- data.frame("n" =  c(),
                            "m" = c(),
                            "estimate" = c())


for (neff in c(250, 500, 1000, 2000)){
  for (meff in c(50, 2000)){
      for (i in 1:1000){
    
      simulation <- toy.example(n = neff, m = meff, output.oracles = F, symetric.po = F, noisier.var.X1 = F, factor.var = 0)
    
      ipsw.pi.oracle <- ipsw.univariate.and.categorical.X(dataframe = simulation, estimand = "ATE", oracle.e = T, oracle.pt = F, oracle.pr = F)
      
      new.row <- data.frame("n" =  c(neff),
                            "m" = c(meff),
                            "estimate" = c(ipsw.pi.oracle))
      
      toy.example.n <- rbind(toy.example.n, new.row)
  
    }
  }
  }


toy.example.n$n <- as.factor(toy.example.n$n )
toy.example.n$m <- as.factor(toy.example.n$m )
```

```{r}
ggplot(toy.example.n, aes(x = n, y = estimate, fill = m)) +
  geom_boxplot(alpha = 0.7) + 
  theme_bw() +
  theme(legend.text = element_text(size = 15, face = "bold"), legend.position = "top", text = element_text(size = 17, face = "bold")) +
  ylab("ATE") +
  xlab("Trial sample's size (n)") +
  facet_wrap(~m, ncol = 2) +
  scale_fill_manual("Target sample size (m)",values=c("orange","red"))
ggsave("./fig/toy_example_2_asympt.pdf", width = 13, height = 5)
```

# Additional covariates: shifted

Reads the output of `toy-example-shifted-covariates.R`.

```{r}
shifted.covariates <- read.csv("./results/toy.example.shifted.covariates.csv")
shifted.covariates$n <- as.factor(shifted.covariates$n)
```

```{r}
additional.covariates.modified <- shifted.covariates %>%
  group_by(covariate.set, pt, pr) %>%
  summarise(variance = var(estimate)) %>%
  pivot_wider(names_from = c(covariate.set), values_from = variance)

additional.covariates.modified$inflation <- additional.covariates.modified$Extended / additional.covariates.modified$Minimal
```


```{r}
additional.covariates.modified$truth <-   (additional.covariates.modified$pt*additional.covariates.modified$pt)/additional.covariates.modified$pr + (1-additional.covariates.modified$pt)*(1-additional.covariates.modified$pt)/(1-additional.covariates.modified$pr)

additional.covariates.modified$borne.sup <- 2*additional.covariates.modified$truth
```


```{r}
ggplot(additional.covariates.modified, aes(x = pt, y = inflation, color = as.factor(pr), group = as.factor(pr))) +
  geom_point(size = 4) +
 geom_line(aes(x = pt, y = truth),  size = 1.9, alpha = 0.6) +
  #geom_line(aes(x = pt, y = borne.sup), color = "darkblue", size = 0.8, alpha = 0.8, linetype = "dashed") +
  theme_minimal() +
  scale_color_manual(expression(p[R](V==1)) ,values=c( "#2A22CD", "#4CCD22")) +
  ylab("Variance inflation") +
  theme(legend.position = "top", text = element_text(size = 12, face = "bold")) +
  xlab(expression(p[T](V==1)))
ggsave("./fig/shifted_covariates.pdf", width = 5, height = 4)
```


# Additional covariates: treatment effect modifier

Now, consider a situation with multiple covariates. Only $X_1$ is necessary, but we have at disposal 3 additional covariates that are treatment effect modifiers but not shifted.

```{r}
big.simulation <- simulation.multivariate.categorical.X(n = 50000, m = 50000, output.oracles = T)
ATE.RCT <- mean(big.simulation[big.simulation$S == 1, "Y_1"]) - mean(big.simulation[big.simulation$S == 1, "Y_0"])
ATE.Target <- mean(big.simulation[big.simulation$S == 0, "Y_1"]) - mean(big.simulation[big.simulation$S == 0, "Y_0"])
```


```{r}
result.adding.covariates <- data.frame("estimate" = c(),
                                       "covariate.set" = c())

for (i in 1:500){
  simulation <- simulation.multivariate.categorical.X(n = 300, m = 1000, output.oracles = F)
 
  simulation$X2 <- as.factor(simulation$X2)
  
  dm <- difference.in.means(simulation, oracle.e.too = F)
 
  tau.hat.minimal <- ipsw.binned(simulation, covariates_names_vector = c("X1"), oracle.e = T, oracle.pt.data = F, oracle.pr = F, outcome_name = "Y", treatment_name = "A")
  
  tau.hat.extended <- ipsw.binned(simulation, covariates_names_vector = c("X1", "X2"), oracle.e = T, oracle.pt.data = F, oracle.pr = F, outcome_name = "Y", treatment_name = "A")
  
 new.row <- data.frame("estimate" = c(dm, tau.hat.minimal, tau.hat.extended),
                                       "covariate.set" = c("DM", "minimal", "extended"))
 
 result.adding.covariates <- rbind(result.adding.covariates, new.row)
}

```

```{r}
result.adding.covariates$estimator.renamed <- case_when(result.adding.covariates$covariate.set == "DM" ~ "Difference-in-means",
                                                        result.adding.covariates$covariate.set == "minimal" ~ "Minimal set",
                                                        result.adding.covariates$covariate.set == "extended" ~ "Extended set")


result.adding.covariates$estimator.renamed <- factor(result.adding.covariates$estimator.renamed , levels = c("Difference-in-means", "Minimal set", "Extended set"))
```


```{r}
ggplot(result.adding.covariates[result.adding.covariates$estimator.renamed != "Difference-in-means",], aes(x =  estimator.renamed, y = estimate, fill = estimator.renamed)) +
  geom_boxplot(alpha = 0.9, size = 0.7) +
  geom_hline(aes(yintercept = ATE.Target, color = "ATE in Target population"),  size = 1, linetype = "dashed", color = "#EA471F") +
  theme_minimal() +
  scale_fill_manual(values=c("#E69F00", "#BCA4F3")) +
  theme(legend.title = element_blank(), legend.text = element_text(size = 13, face = "bold"), legend.position = "none", axis.text = element_text(size = 13, face = "bold", color = "black")) +
  ylab("ATE") +
  xlab("") 
ggsave("./fig/toy_precision_covariates.pdf", width = 4.3, height = 4.8)
```


# Asymptotics regimes

Reads output of `finite-sample-toy-example.R`.

```{r}
asymptotic.regimes <- read.csv("./results/finite.sample.toy.example.csv")
names(asymptotic.regimes)[names(asymptotic.regimes) == "method"] <- "Regime"
```


```{r}
asymptotic.regimes$Regime <- case_when(asymptotic.regimes$Regime == "IPSW: m = m" ~ "IPSW: m = n",
                                       asymptotic.regimes$Regime == "IPSW: n = √m" ~ "IPSW: m = n*n",
                                       asymptotic.regimes$Regime == "IPSW: n = 2m" ~ "IPSW: m = n/2",
                                       asymptotic.regimes$Regime == "IPSW: n = m*m" ~ "IPSW: m = √n",
                                       asymptotic.regimes$Regime == "IPSW: n = m/2" ~ "IPSW: m = 2n",
                                       asymptotic.regimes$Regime  %in% c("Semi-oracle", "Completely-oracle") ~ asymptotic.regimes$Regime)
```


```{r}
asymptotic.regimes$IPSW <- ifelse(asymptotic.regimes$Regime == "Completely-oracle" | asymptotic.regimes$Regime == "Semi-oracle", "Completely or semi oracle", "Estimated")
```


```{r}
risks <- asymptotic.regimes %>%
  group_by(n,m, Regime, IPSW) %>%
  summarise(Bias = mean(estimate) - TARGET.ATE, Variance =  var(estimate), Risk = Bias*Bias + Variance)
```


```{r}
library(paletteer)
ggplot(risks, aes(x = n, y = Risk, color = Regime, linetype = IPSW)) +
  geom_line(size = 1.5) +
  geom_point(size = 4) +
  theme_minimal() +
  ylab(expression(paste( "Quadratic error:  ", E((hat(tau)-tau)**2)))) +
  scale_color_paletteer_d("ggsci::uniform_startrek") +
   theme(legend.text = element_text(size = 8, face = "bold"), axis.text = element_text(size = 8, face = "bold", color = "black"), legend.position =  "none") +
  xlab("Trial sample size (n)")
ggsave("./fig/toy.example.risks.png", width = 4.5, height = 3.7)
```



```{r}
asymptotic.regimes.estimated <- asymptotic.regimes[asymptotic.regimes$IPSW == "Estimated",]%>%
  group_by(n, m, Regime, IPSW) %>%
  summarise(Variance = var(estimate)*min(n,m))

asymptotic.regimes.oracle <- asymptotic.regimes[asymptotic.regimes$IPSW != "Estimated",]%>%
  group_by(n, Regime, IPSW) %>%
  summarise(Variance = var(estimate)*n)


asymptotic.regimes <- rbind(asymptotic.regimes.estimated, asymptotic.regimes.oracle) 
```

```{r}

ggplot(asymptotic.regimes[asymptotic.regimes$n <300,], aes(x = n, y = Variance, color = Regime, linetype = IPSW)) +
  geom_line(size = 1.5) +
  geom_point(size = 4) +
  theme_minimal() +
  ylab(expression(paste("min(n,m) ",Var(hat(tau))))) +
  scale_color_paletteer_d("ggsci::uniform_startrek") +
   theme(legend.text = element_text(size = 8, face = "bold"), axis.text = element_text(size = 8, face = "bold", color = "black"), legend.position = "left") +
  xlab("Trial sample size (n)")
ggsave("./fig/toy.example.regimes.png", width = 4.5, height = 3.5)
```

